---
title: "cMD metadata dictionary formatting for Genestack ODM"
author: "Sehyun Oh"
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{cMD dictionaries for Genestack}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    number_sections: yes
    toc: yes
    toc_depth: 4
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(comment = "#>", 
                      collapse = TRUE, 
                      message = FALSE, 
                      warning = FALSE, 
                      eval = FALSE)
```

```{r}
dd <- readr::read_csv("~/OmicsMLRepo/OmicsMLRepoData/inst/extdata/cMD_data_dictionary.csv")
```

# NCIT
```{bash eval=FALSE}
## Download `ncit.obo` in `~/Projects/curatedMetagenomicDataODM` directory
wget https://purl.obolibrary.org/obo/ncit.obo
```

## recist
```{r echo=FALSE, eval=FALSE}
ncit <- ontologyIndex::get_ontology("~/Projects/curatedMetagenomicDataODM/ncit.obo")
recist_terms <- c("NCIT:C159715", "NCIT:C159547", "NCIT:C159716", "NCIT:C159546")
writeLines(recist_terms, "dictionaries_terms/recist.txt")
recist_ind <- which(ncit$id %in% recist_terms)
recist_obo <- lapply(ncit, function(x) {x[recist_ind]})

library(BiocSet)
x <- import("~/Projects/curatedMetagenomicDataODM/ncit.obo")
y <- x %>% filter(element %in% recist_terms)
export(y, "dictionaries/cmd_recist.obo")
```

```{r}
recist_terms <- c("NCIT:C159715", "NCIT:C159547", "NCIT:C159716", "NCIT:C159546")
writeLines(recist_terms, "dictionaries_terms/recist.txt")
```

```{bash eval=FALSE}
robot extract --method subset \
    --input ontologies/ncit.obo \
    --term-file dictionaries_terms/recist.txt \
    --output dictionaries_owl/cMD_disease_response_recist.owl
```

## age_group
```{r}
attrName <- "age_group"
attrVals <- dd %>% filter(col.name == attrName) %>%
    select(ontology) %>% unlist %>%
    strsplit(., split = ";") %>% .[[1]] %>% .[c(5,3,1,2,4)]
unique(OmicsMLRepoR::get_ontologies(attrVals)) # ontology used for this attribute

## Save the list of static ontology terms
fpath <- paste0("dictionaries_terms/", attrName,".txt")
writeLines(attrVals, fpath)
```

```{bash}
attrName="age_group"
robot extract --method subset \
    --input ontologies/ncit.obo \
    --term-file dictionaries_terms/${attrName}.txt \
    --output dictionaries_owl/cmd_${attrName}.owl
```

## age_unit
```{r}
attrName <- "age_unit"
attrVals <- dd %>% filter(col.name == attrName) %>%
    select(ontology) %>% unlist %>%
    strsplit(., split = ";") %>% .[[1]]
unique(OmicsMLRepoR::get_ontologies(attrVals)) # ontology used for this attribute

## Save the list of static ontology terms
fpath <- paste0("dictionaries_terms/", attrName,".txt")
writeLines(attrVals, fpath)
```

```{bash}
attrName="age_unit"
robot extract --method subset \
    --input ontologies/ncit.obo \
    --term-file dictionaries_terms/${attrName}.txt \
    --output dictionaries_owl/cmd_${attrName}.owl
```

## control
```{r}
attrName <- "control"
attrVals <- dd %>% filter(col.name == attrName) %>%
    select(ontology) %>% unlist %>%
    strsplit(., split = ";") %>% .[[1]]
unique(OmicsMLRepoR::get_ontologies(attrVals)) # ontology used for this attribute

## Save the list of static ontology terms
fpath <- paste0("dictionaries_terms/", attrName,".txt")
writeLines(attrVals, fpath)
```

```{bash}
attrName="control"
robot extract --method subset \
    --input ontologies/ncit.obo \
    --term-file dictionaries_terms/${attrName}.txt \
    --output dictionaries_owl/cmd_${attrName}.owl
```

## neonatal_delivery_procedure
```{r}
attrName <- "neonatal_delivery_procedure"
attrVals <- dd %>% filter(col.name == attrName) %>%
    select(ontology) %>% unlist %>%
    strsplit(., split = ";") %>% .[[1]]
unique(OmicsMLRepoR::get_ontologies(attrVals)) # ontology used for this attribute

## Save the list of static ontology terms
fpath <- paste0("dictionaries_terms/", attrName,".txt")
writeLines(attrVals, fpath)
```

```{bash}
attrName="neonatal_delivery_procedure"
robot extract --method subset \
    --input ontologies/ncit.obo \
    --term-file dictionaries_terms/${attrName}.txt \
    --output dictionaries_owl/cmd_${attrName}.owl
```




## treatment (dynamic enums)
```{bash}
## Used dynamic enum node for cMD_treatment
robot extract --method MIREOT \
    --input ontologies/ncit.obo \
    --branch-from-term "NCIT:C1908" \
    --output dictionaries_owl/cmd_treatment_enums.owl
```

## country (dynamic enums)
```{bash}
attrName="country"
robot extract --method MIREOT \
    --input ontologies/ncit.obo \
    --branch-from-term "NCIT:C25464" \
    --output dictionaries_owl/cmd_${attrName}_enums.owl
```

## obgyn_menopause (dynamic enums)
```{bash}
attrName="obgyn_menopause"
robot extract --method MIREOT \
    --input ontologies/ncit.obo \
    --branch-from-term "NCIT:C106541" \
    --output dictionaries_owl/cmd_${attrName}_enums.owl
```

## disease (dynamic enums)
```{r}
obo <- ontologyIndex::get_ontology("~/Projects/curatedMetagenomicDataODM/ontologies/efo.obo")
```

```{bash}
attrName="disease"
robot extract --method MIREOT \
    --input ontologies/ncit.obo \
    --branch-from-term "NCIT:C7057" \
    --output dictionaries_owl/cmd_${attrName}_NCIT_enums.owl
    
## Download `efo.obo` from http://www.ebi.ac.uk/efo/efo.obo
## wget http://www.ebi.ac.uk/efo/efo.obo in the `ontologies` directory
robot extract --method MIREOT \
    --input ontologies/efo.obo \
    --branch-from-term "http://www.ebi.ac.uk/efo/EFO_0000408" \
    --output dictionaries_owl/cmd_${attrName}_EFO_enums.owl

robot merge \
    --input dictionaries_owl/cmd_disease_NCIT_enums.owl \
    --input dictionaries_owl/cmd_disease_EFO_enums.owl \
    --output dictionaries_owl/cmd_${attrName}_enums.owl
```




# MRO
Download `mro.owl` from https://obofoundry.org/.

## hla (dynamic enums)
```{bash}
attrName="hla"
robot extract --method MIREOT \
    --input ontologies/mro.owl \
    --branch-from-term "MRO:0001676" \
    --output dictionaries_owl/cmd_${attrName}_enums.owl
```

# FMA
Download `fma.owl` version 5.0.0 from https://bioportal.bioontology.org/ontologies/FMA.





# Simple CSV
## smoker
`snomed.owl` is a bit tricky to download. I made the simple csv for this attribute.
```{r}
attrName <- "smoker"
attrVals <- dd %>% filter(col.name == attrName) %>%
    select(allowedvalues) %>% unlist %>%
    strsplit(., split = ";") %>% .[[1]] %>% c("label", .)

## Save the list of static ontology terms
fpath <- paste0("dictionaries_owl/cmd_", attrName,".csv")
writeLines(attrVals, fpath)
```

## dietary_restriction
No ontology terms used for this attribute.
```{r}
attrName <- "dietary_restriction"
attrVals <- dd %>% filter(col.name == attrName) %>%
    select(allowedvalues) %>% unlist %>%
    strsplit(., split = ";") %>% .[[1]] %>% c("label", .)

## Save the list of static ontology terms
fpath <- paste0("dictionaries_owl/cmd_", attrName,".csv")
writeLines(attrVals, fpath)
```

## curator
No ontology terms used for this attribute.
```{r}
attrName <- "curator"
attrVals <- dd %>% filter(col.name == attrName) %>%
    select(allowedvalues) %>% unlist %>%
    strsplit(., split = ";") %>% .[[1]] %>% 
    unique 
updatedAttrVals <- c(attrVals, "Alice_Bassetto") %>% sort %>% c("label", .)


## Save the list of static ontology terms
fpath <- paste0("dictionaries_owl/cmd_", attrName,".csv")
writeLines(updatedAttrVals, fpath)
```

## fmt_role
No ontology terms used for this attribute.
```{r}
attrName <- "fmt_role"
attrVals <- dd %>% filter(col.name == attrName) %>%
    select(allowedvalues) %>% unlist %>%
    strsplit(., split = ";") %>% .[[1]] %>% c("label", .)

## Save the list of static ontology terms
fpath <- paste0("dictionaries_owl/cmd_", attrName,".csv")
writeLines(attrVals, fpath)
```

## lifestyle
No ontology terms used for this attribute.
```{r}
attrName <- "lifestyle"
attrVals <- dd %>% filter(col.name == attrName) %>%
    select(allowedvalues) %>% unlist %>%
    strsplit(., split = ";") %>% .[[1]] %>% c("label", .)

## Save the list of static ontology terms
fpath <- paste0("dictionaries_owl/cmd_", attrName,".csv")
writeLines(attrVals, fpath)
```

## neonatal_feeding_method
No ontology terms used for this attribute.
```{r}
attrName <- "neonatal_feeding_method"
attrVals <- dd %>% filter(col.name == attrName) %>%
    select(allowedvalues) %>% unlist %>%
    strsplit(., split = ";") %>% .[[1]] %>% c("label", .)

## Save the list of static ontology terms
fpath <- paste0("dictionaries_owl/cmd_", attrName,".csv")
writeLines(attrVals, fpath)
```

## tumor_staging_ajcc
No ontology terms used for this attribute.
```{r}
attrName <- "tumor_staging_ajcc"
attrVals <- dd %>% filter(col.name == attrName) %>%
    select(allowedvalues) %>% unlist %>%
    strsplit(., split = ";") %>% .[[1]] %>% c("label", .)

## Save the list of static ontology terms
fpath <- paste0("dictionaries_owl/cmd_", attrName,".csv")
writeLines(attrVals, fpath)
```

## zygosity
No ontology terms used for this attribute.
```{r}
attrName <- "zygosity"
attrVals <- dd %>% filter(col.name == attrName) %>%
    select(allowedvalues) %>% unlist %>%
    strsplit(., split = ";") %>% .[[1]] %>% c("label", .)

## Save the list of static ontology terms
fpath <- paste0("dictionaries_owl/cmd_", attrName,".csv")
writeLines(attrVals, fpath)
```

## family_role
No ontology terms used for this attribute.
```{r}
attrName <- "family_role"
attrVals <- dd %>% filter(col.name == attrName) %>%
    select(allowedvalues) %>% unlist %>%
    strsplit(., split = ";") %>% .[[1]] %>% c("label", .)

## Save the list of static ontology terms
fpath <- paste0("dictionaries_owl/cmd_", attrName,".csv")
writeLines(attrVals, fpath)
```

## feces_phenotype_metric
Contains values from different ontologies, including SNOMED.
```{r}
attrName <- "feces_phenotype_metric"
attrVals <- dd %>% filter(col.name == attrName) %>%
    select(allowedvalues) %>% unlist %>%
    strsplit(., split = ";") %>% .[[1]] %>% c("label", .)

## Save the list of static ontology terms
fpath <- paste0("dictionaries_owl/cmd_", attrName,".csv")
writeLines(attrVals, fpath)
```

## obgyn_pregnancy
Straight-foward concepts
```{r}
attrName <- "obgyn_pregnancy"
attrVals <- dd %>% filter(col.name == attrName) %>%
    select(allowedvalues) %>% unlist %>%
    strsplit(., split = ";") %>% .[[1]] %>% c("label", .)

## Save the list of static ontology terms
fpath <- paste0("dictionaries_owl/cmd_", attrName,".csv")
writeLines(attrVals, fpath)
```

## sex
```{r}
attrName <- "sex"
attrVals <- dd %>% filter(col.name == attrName) %>%
    select(allowedvalues) %>% unlist %>%
    strsplit(., split = ";") %>% .[[1]] %>% c("label", .)

## Save the list of static ontology terms
fpath <- paste0("dictionaries_owl/cmd_", attrName,".csv")
writeLines(attrVals, fpath)
```

## probing_pocket_depth
Can't find the correct version of `fma.owl`
```{r}
attrName <- "probing_pocket_depth"
attrVals <- dd %>% filter(col.name == attrName) %>%
    select(allowedvalues) %>% unlist %>%
    strsplit(., split = ";") %>% .[[1]] %>% c("label", .)

## Save the list of static ontology terms
fpath <- paste0("dictionaries_owl/cmd_", attrName,".csv")
writeLines(attrVals, fpath)
```

## dna_extraction_kit
```{r}
attrName <- "dna_extraction_kit"
attrVals <- dd %>% filter(col.name == attrName) %>%
    select(allowedvalues) %>% unlist %>%
    strsplit(., split = ";") %>% .[[1]] %>% c("label", .)

## Save the list of static ontology terms
fpath <- paste0("dictionaries_owl/cmd_", attrName,".csv")
writeLines(attrVals, fpath)
```
